Biden More Favorable Among Female, Black, Asian, Hispanic, Middle Eastern, and Non-gun Owning Voters in the 2020 US Presidential Election

### 2024-03-16 paper/paper.qmd
Update paper with citations

---
title: "Biden More Favorable Among Female, Black, Asian, Hispanic, Middle Eastern, and Non-gun Owning Voters in the 2020 US Presidential Election"
author: 
  - Tam Ly
  - Renfrew Ao-Ieong
  - Rahma Binth Mohammad
thanks: "Code and data are available at: [https://github.com/atn-ly/political_support_in_the_us](https://github.com/atn-ly/political_support_in_the_us)"
date: today
date-format: long
abstract: "The outcome of the 2020 US Presidential Election served as a reflection of the beliefs and attitudes held by Americans. Using data obtained during the fall of 2020 from the Cooperative Election Study, we investigated if a person is likely to vote for Biden, based only on knowing their gender, race, and gun ownership status. The results of our analysis using a Bayesian logistic regression model show that female, Black, Asian, Hispanic, Middle Eastern, and non-gun owning individuals are more likely to support Biden. Our discovery highlights the importance of women’s issues, racial justice, and gun policy on American voters."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

The United States (US) presidential election of 2020 resulted in President Joe Biden’s victory, making him the 46th President of the US. The two top runners were Biden for the Democrats and former President Donald Trump for the Republicans. The presidential race between the two was of significant interest to statisticians and polling experts due to its potential impact on polling models and election forecasting. To provide some background, President Biden ran on his merit as a former Vice President of eight years and had faced lots of criticism during his time in office. Moreover, Former President Trump had controversial and career-damaging events. During the 2016 elections, Trump even faced accusations regarding election rigging (cite). Before the 2020 elections, Trump was charged with misconduct while in office(cite). However, the senate acquitted him of those misconduct charges. Thus, both Biden and Trump faced have previously criticism from the electorate. An interesting question to consider is despite these, what characteristics did the voters have? Do they tell us a story on the type of people that voted for Biden and the type that voted for Trump? Answering these questions can help develop polling models and future election forecasting.

@@

Based on our model and analysis, we found that certain demographic groups were mor include to support either candidate. Our findings revealed that women were more likely to vote for Biden compared to Trump. This finding is consistent with broader trends in political behaviour as gender has previously shown to influence voting preferences (cite). Biden’s democratic policies were more convincing to women and will be discussed in @sec-discussion. Another major finding was that Black, Asian, Hispanic, and Middle Eastern individuals are also more likely to vote for Biden. Biden has been widely known to run on a more progressive platform, giving importance on racial justice and addressing racial disparities. Our finding emphasizes that people from diverse and ethnic backgrounds are more inclined to vote for progressive representatives. Our final major finding was that non-gun owners are also more likely to vote for Biden. Gun owners often have strong opinions on gun control policies. Our finding shows that those who do not possess guns favored Biden’s gun control and other firearm-related policies, as opposed to Trump’s policies.

This paper was structured to communicate our mode, results, and analysis to the reader in a strategic manner. Data for this analysis and the different applications used will be further introduced in @sec-data. @sec-model provides the model set-up and the justification for the use of that model. @sec-results will show results and @sec-discussion is a discussion of the results including the paper’s weaknesses and biases.

@@

The survey data for this paper is from the Harvard Dataverse Repository. The Cooperative Election Study is a sample survey in the Harvard Dataverse Repository that consists of pre-election and post-election questions with 61000 answers. They provide a guide and a dataset for numerous years. It is also open to the public. For this paper, we utilized the data from the 2020 Presidential Elections(cite). The study was done to gain an understanding of “how Americans view Congress and hold their representatives accountable during elections, how they voted and their electoral experiences, and how their behaviour and experiences vary with political geography and social context” (cite). The outcome of the 2020 election and every presidential election usually changes the direction of the country and demonstrates the perspectives of the majority of the electorate. Thus, this dataset was able to give us some clarity on perceptions of the electorate. There were similar datasets available, but we chose this one due to relevance to our research question. The questionnaire asked various questions on race, political party preference, voting preference, ownership of a gun and have numerous multiple choice answers available for each question. It was also easily accessible and we were able to adapt and adjust code from Telling Stories with Data to suit our needs [@rohan].

@@

In this paper, we measured key variables related to voter behaviour. From the dataset, we selected `gender`, `race`, `gunown`. In Figure 1, we can see support for candidate based on gender and race. For gender, we considered two options, male or female. For race, we considered 8 options, White, Black, Hispanic, Asian, Native American, Middle Eastern, two or more races, and other. Note that only one selection was allowed in the original study for both gender and race. Figure 2 illustrates the support for each candidate based on gun ownership status, with four options. Individuals who personally own a gun, who don’t personally own one, but someone in their household owns a gun, no one in the household owns a gun, and not sure. We also selected the variable that recorded whether an individual voted for Biden or Trump in the 2020 Presidential election, `CC20_410`. This allowed us to establish a binary outcome variable for the probability of voting Biden or Trump in our model in @sec-model. The variables measured here allowed us to identify variation in political attitudes and voting patterns.

Data was downloaded and cleaned to suit our needs and during the data cleaning process, we made sure to only include participants who were adults registered to vote. We also removed rows with NA values for `gunown`. This was due to the fact that we wanted to see individuals who answered all the questions for our variables of interest. We almost limited the participants to those who explicitly stated they voted for Biden or Trump. Other options were available to participants such as “I did not vote” and “not sure”.

>>>>

---
title: "Biden More Favorable Among Female, Black, Asian, Hispanic, Middle Eastern, and Non-gun Owning Voters in the 2020 US Presidential Election"
author: 
  - Tam Ly
  - Renfrew Ao-Ieong
  - Rahma Binth Mohammad
thanks: "Code and data are available at: [https://github.com/atn-ly/political_support_in_the_us](https://github.com/atn-ly/political_support_in_the_us)"
date: 16 March 2024
date-format: long
abstract: "The outcome of the 2020 US Presidential Election served as a reflection of the beliefs and attitudes held by Americans. Using data obtained during the fall of 2020 from the Cooperative Election Study, we investigated if a person is likely to vote for Biden, based only on knowing their gender, race, and gun ownership status. The results of our analysis using a Bayesian logistic regression model show that female, Black, Asian, Hispanic, Middle Eastern, and non-gun owning individuals are more likely to support Biden. Our discovery highlights the importance of women’s issues, racial justice, and gun policy on American voters."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

The United States (US) presidential election of 2020 resulted in President Joe Biden’s victory, making him the 46th President of the US. The two top runners were Biden for the Democrats and former President Donald Trump for the Republicans. The presidential race between the two was of significant interest to statisticians and polling experts due to its potential impact on polling models and election forecasting. To provide some background, President Biden ran on his merit as a former Vice President of eight years and had faced lots of criticism during his time in office. Moreover, Former President Trump had controversial and career-damaging events. Before the 2020 elections, Trump was charged with misconduct while in office but the senate acquitted him of those misconduct charges [@impeachment]. Thus, both Biden and Trump faced have previously criticism from the electorate. An interesting question to consider is despite these, what characteristics did the voters have? Do they tell us a story on the type of people that voted for Biden and the type that voted for Trump? Answering these questions can help develop polling models and future election forecasting.

@@

Based on our model and analysis, we found that certain demographic groups were more include to support either candidate. Our findings revealed that women were more likely to vote for Biden compared to Trump. This finding is consistent with broader trends in political behavior, as gender has previously shown to influence voting preferences. Biden’s democratic policies were more convincing to women and will be discussed in @sec-discussion. Another major finding was that Black, Asian, Hispanic, and Middle Eastern individuals are also more likely to vote for Biden. Biden has been widely known to run on a more progressive platform, giving importance on racial justice and addressing racial disparities. Our finding emphasizes that people from diverse and ethnic backgrounds are more inclined to vote for progressive representatives. Our final major finding was that non-gun owners are also more likely to vote for Biden. Gun owners often have strong opinions on gun control policies. Our finding shows that those who do not possess guns favored Biden’s gun control and other firearm-related policies, as opposed to Trump’s policies.

This paper was structured to communicate our model, results, and analysis to the reader in a strategic manner. Data for this analysis and the different applications used will be further introduced in @sec-data. @sec-model provides the model set-up and the justification for the use of that model. @sec-results will show results and @sec-discussion is a discussion of the results including the paper’s weaknesses and biases.

@@

The survey data for this paper is from the Harvard Dataverse Repository. The Cooperative Election Study is a sample survey in the Harvard Dataverse Repository that consists of pre-election and post-election questions with 61000 answers. They provide a guide and a dataset for numerous years. It is also open to the public. For this paper, we utilized the data from the 2020 Presidential Elections [@CES]. The study was done to gain an understanding of “how Americans view Congress and hold their representatives accountable during elections, how they voted and their electoral experiences, and how their behaviour and experiences vary with political geography and social context” [@CES]. The outcome of the 2020 election and every presidential election usually changes the direction of the country and demonstrates the perspectives of the majority of the electorate. Thus, this dataset was able to give us some clarity on perceptions of the electorate. There were similar datasets available, but we chose this one due to relevance to our research question. The questionnaire asked various questions on race, political party preference, voting preference, ownership of a gun and have numerous multiple choice answers available for each question. It was also easily accessible and we were able to adapt and adjust code from Telling Stories with Data [@rohan] to suit our needs.

@@

In this paper, we measured key variables related to voter behaviour. From the dataset, we selected `gender`, `race`, `gunown`. In Figure 1, we can see support for candidate based on gender and race. For gender, we considered two options, male or female. For race, we considered eight options, White, Black, Hispanic, Asian, Native American, Middle Eastern, two or more races, and other. Note that only one selection was allowed in the original study for both gender and race. Figure 2 illustrates the support for each candidate based on gun ownership status, with four options. Individuals who personally own a gun, who don’t personally own one, but someone in their household owns a gun, no one in the household owns a gun, and not sure. We also selected the variable that recorded whether an individual voted for Biden or Trump in the 2020 Presidential election, `CC20_410`. This allowed us to establish a binary outcome variable for the probability of voting Biden or Trump in our model in @sec-model. The variables measured here allowed us to identify variation in political attitudes and voting patterns.

Data was downloaded and cleaned to suit our needs and during the data cleaning process, we made sure to only include participants who were adults registered to vote. We also removed rows with NA values for `gunown`. This was due to the fact that we wanted to see individuals who answered all the questions for our variables of interest. We also limited the participants to those who explicitly stated they voted for Biden or Trump. Initially, other options were available to participants such as “I did not vote” and “not sure”.
